---
title: "Peer Assessment 1"
author: "Igor Tomashevskiy"
output:
  html_document:
    fig_width: 8
    keep_md: yes
  pdf_document: default
---
The dataset is stored in a comma-separated-value (CSV) file: activity.csv
The variables included in this dataset -  
steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
date: The date on which the measurement was taken in YYYY-MM-DD format  
interval: Identifier for the 5-minute interval in which measurement was taken.    

**Data pre-processing**  

```{r}
pa_data<-read.csv("activity.csv",colClasses=c("integer",'Date','integer'),na.strings="NA")
summary(pa_data)
```
There are 17,568 observations in this dataset. Number of records with missing values is 2304.  

**Mean total number of steps taken per day**  

New data set pa_data_agg will be created to calculate the mean and median of the total number of steps taken per day
```{r}
pa_data_agg<-aggregate(steps~date,pa_data,sum)
mean(pa_data_agg$steps)
median(pa_data_agg$steps)
```
Mean Total Steps =10766  
Median Total Steps = 10765  


We will use the ggplot function to make a histogram of the total number of steps taken each day.

```{r}
library(ggplot2)
h1<-ggplot(pa_data,aes(date,steps))+geom_bar(stat="identity",binwidth=1,color='blue')+labs(title="Histogram of the total number of steps taken each day",x="Date",y="Steps")
print(h1)
```

**Average daily activity pattern**  

To answer the question about the average daily activity pattern we need to modify the dataset:  

```{r}
daily_activity<-aggregate(steps~interval,pa_data,mean)
```
The code to produce the a time series plot:  

```{r}
t1<-ggplot(daily_activity,aes(interval,steps))+geom_line(color="red",size=1)+labs(y="Average Steps")
print(t1)
````

Next we will calculate the 5-minute interval that contains the maximum number of steps on average across all the days
in the dataset.  

```{r}
interval<-daily_activity[rev(order(daily_activity$steps)),]
head(interval,1)
````

The 5 min interval is 835  

**Imputing missing values**  

There are a number of days/intervals with missing values in the original dataset. The presence of missing days may introduce bias into some calculations or summaries of the data.  
The md.pattern() function in the mice package produces a tabulation of the missing data patterns.

```{r}
library(mice)
md.pattern(pa_data)
        
```
The second row describes the pattern with 'no missing values' except for steps. There are 2304 rows missing steps alone.
It is also easy to discern patterns visually.
```{r warning=FALSE}
library(VIM)
aggr(pa_data, prop=FALSE, numbers=TRUE)
```

A quick fix for the missing data is to replace them by the mean. However, it will underestimate the variance, disturb the relations between variables, bias all estimates other than the mean (will bias mean estimate also if data are not 'misisng completely at random'). Mean imputation could perhaps only be used as a fix when we have only a few values missing. Our data set contains about 13% of records with missing data.
Better approach is to use mice package in R. By default, each varible with missing values is predicted from all other variables in the dataset. We will use the simple form of mice function with one dataset (m=1). In practice m is taken larger.

```{r}
impute_data<-pa_data
impute_data$num_date<-as.numeric(pa_data$date)
imp<-mice(impute_data[,c(1,3,4)],m=1,seed=123)
temp_data<-complete(imp)
#colnames(temp_data)[1]<-"imp_steps"
impute_data<-cbind(impute_data,temp_data$steps)
colnames(impute_data)[5]<-"imp_steps"
````
```{r}
h2<-ggplot(impute_data,aes(date,imp_steps))+geom_bar(stat="identity",binwidth=1,color='blue')+labs(title="Histogram of the total number of steps taken each day",x="Date",y="Steps")
print(h2)
```
The histogram is quite similar to the first histogram.  

```{r}
pa_data_agg<-aggregate(imp_steps~date,impute_data,sum)
mean(pa_data_agg$imp_steps)
median(pa_data_agg$imp_steps)

````
**Activity patterns between weekdays and weekends**  
```{r}
library(lattice)
impute_data$day<-weekdays(impute_data$date)
impute_data$day[impute_data$day=="Saturday"|impute_data$day=="Sunday"]<-'weekend'
impute_data$day[impute_data$day!='weekend']<-'weekday'
impute_data$day<-as.factor(impute_data$day)
impute_data_agg<-aggregate(imp_steps~interval+day,data=impute_data,mean)
xyplot(imp_steps~interval|day,impute_data_agg,type='l',layout=c(1,2),ylab="steps")
```





